Faithful Variable Screening for High-Dimensional Convex Regression

### 2014-07-21 tex/scam/additive_faithful.tex
edited through end of section.  important: commented out proposition 3.1, as it looks like an old version of the theorem?

\begin{example}
\label{examp:gaussian_counterexample}
Consider a quadratic form $f( \mathbf{x}) = \mathbf{x}^\tran H \mathbf{x} + c^\tran \mathbf{x}$ and a Gaussian distribution $X \sim N(0, \Sigma)$ where $\Sigma_{jj} = 1$ for all $j$.
As we show in the appendix, the additive approximation has the
following closed form.
\begin{enumerate}
\item If $f$ does not depend on $x_j$, then $f^*_j(x_j) = 0$.
\item If $f$ depends on $x_j$, then
\begin{equation}
f^*_j(x_j) = H_j^\tran \Sigma_j x_j^2 + c_j x_j
\end{equation}
where $H_j, \Sigma_j$ are the $j$-th rows of $H_j, \Sigma_j$, respectively. 
\end{enumerate}
Let $H = \begin{pmatrix}1 & 2 \\ 2 & 5\end{pmatrix}$ and let $\Sigma
= \begin{pmatrix}1 & -\alpha \\ -\alpha & 1\end{pmatrix}$. 
It can be checked that if if $\alpha=\frac{1}{2}$, then $f^*_1 = 0$
and additive faithfulness is violated.  Moreover, if $\alpha > \frac{1}{2}$, then $f^*_1$ is a concave function.
\end{example}

@@

Although convex functions are additively faithful, it is difficult to estimate the optimal additive functions $f^*_k$'s (as defined in equation~\ref{eqn:unconstrained_additive}) because $f^*_k$ need not be a convex function, as example~\ref{examp:gaussian_counterexample} shows. $f^*_k$ can be shown to be smooth but it is preferable to use shape-constraints whenever possible to avoid the introduction of the additional smoothing bandwidth.

Since the true regression function $f$ is convex, it is natural to ask when it is sufficient to estimate an convex additive model:
\begin{equation}
\label{eqn:convex_additive}
\{ f^*_k \}_{k=1}^p = \arg\min \Big \{ 
    \E\big( f(X) - \sum_{k=1}^p f_k(X_k) \big)^2 \,:\, f_k \in \mathcal{C}^1, \, \E f_k(X_k) = 0 \Big \}
\end{equation}
where $\mathcal{C}^1$ is the set of univariate convex functions.

The convex functions $f^*_k$'s are not additively faithful by themselves unfortunately. But, faithfulness can be restored by coupling the $f^*_k$'s with a set of concave functions:
\begin{equation}
\label{eqn:concave_postprocess}
g^*_k = \arg\min \Big\{
   \E\big( f(X) - \sum_{k' \neq k} f^*_{k'}(X_{k'}) - g_k \big)^2 
    \,:\, g_k \in \mh \mathcal{C}^1, \E g_k(X_k) = 0 
  \Big\}
\end{equation}

\begin{theorem}
\label{thm:acdc_faithful}
Suppose $p(\mathbf{x})$ is a positive density on $C=[0,1]^p$ and satisfies the boundary points condition. Suppose that $f$ is convex and twice-differentiable.

Suppose that $\partial_{x_k} f$, $\partial_{x_k} p( \mathbf{x}_{-k} \given x_k )$, and $\partial_{x_k}^2 p( \mathbf{x}_{-k} \given x_k)$ are all continuous as functions on $C$.\\

Then, $f^*_k = 0$ and $g^*_k = 0$ (as defined in equation~\ref{eqn:convex_additive} and equation~\ref{eqn:concave_postprocess}) implies that $f$ does not depend on $x_k$, i.e., $\partial_{x_k} f(\mathbf{x}) = 0$ with probability 1.
\end{theorem}

Theorem~\ref{thm:acdc_faithful} suggests a two-stage screening procedure for variable selection. We first fit a sparse \emph{convex} additive model and then, for every variable marked as irrelevant in the first stage, we allow ourself to change our mind by fitting an univariate \emph{concave} function. We refer to this procedure as AC/DC (additive convex / decoupled concave) and describe it in more details in section~\ref{sec:acdc}.

\begin{proof} (of theorem~\ref{thm:acdc_faithful}) \\
Fix $k$. Let $f^*_k, g^*_k$'s be defined as equation~\ref{eqn:convex_additive} and equation~\ref{eqn:concave_postprocess}.

By definition of $f^*_k$ and $g^*_k$, we have that:
\begin{align*}
f^*_k &= \arg\min_{f_k} \Big\{
   \E\big( f(X) - \sum_{k' \neq k} f^*_{k'}(X_{k'}) - g_k \big)^2 
    \,:\, f_k \in  \mathcal{C}^1,\, \E f_k(X_k) = 0 
  \Big\} \\
g^*_k &= \arg\min_{g_k} \Big\{
   \E\big( f(X) - \sum_{k' \neq k} f^*_{k'}(X_{k'}) - g_k \big)^2 
    \,:\, g_k \in \mh \mathcal{C}^1,\, \E g_k(X_k) = 0 
  \Big\}
\end{align*}

Therefore, $f^*_k = 0$ and $f^*_k = 0$ implies that $\E[ f(X) - \sum_{k'\neq k} f^*_{k'}(X_{k'}) \given x_k] = 0$.

Now we use corollary~\ref{cor:faithfulness_extension}. We let $\phi(\mathbf{x}_{-k}) = f(\mathbf{x}) - \sum_{k' \neq k} f^*_{k'} (x_{k'})$ and conclude that $f$ does not depend on $x_k$. 
\end{proof}

\begin{corollary} 
\label{cor:faithfulness_extension}
Suppose $p(x)$ is a positive density on $[0,1]^p$ and it satisfies the boundary points condition.

For any function $\phi(X_{-k})$ that does not depend on $X_k$:
 \begin{align*}
f^*_k(x_k) &= \arg\min_{f_k} \mathbb{E} \Big( f(X) 
           - \phi(X_{-k}) - f_k(X_k) \Big)^2 
      = \mathbb{E}\Big[ f(X) - \phi(X_{-k}) \,|\, x_k\Big] 
\end{align*}
We have that $ f^*_k = 0 \Rightarrow \partial_{x_k} f(x) = 0$.
\end{corollary}

\begin{proof}
In the proof of theorem~\ref{thm:convex_faithful}, the only property of $r(\mathbf{x}_{-k})$ we used was the fact that $\partial_{x_k} r(\mathbf{x}_{-k}) = 0$. Therefore, the proof here is identical to that of theorem~\ref{thm:convex_faithful} except we let $\phi(\mathbf{x}_{-k})= r(\mathbf{x}_{-k}).$
\end{proof}

\begin{proposition}
\label{prop:shape_approx_univariate}
Let $C \subset \R^p$ be a compact set and let $h : C \rightarrow \R$. Let $p(x)$ be a positive density on $C$ and suppose $\E h(X) = 0$. Suppose that $\partial_{x_k} h(x)$, $\partial_{x_k} p( x \given x_k)$, and $\partial^2_{x_k} p( x \given x_k )$ are all continuous as functions on $C$. Suppose that $\partial^2_{x_k} h(x) \geq 0$.

Let 
\begin{align*}
f^*_k(x_k) &= \arg\min_{f_k} \Big\{
   \E \Big( h(X) - f_k(X_k) \Big)^2 \,:\, f_k \in \mathcal{C}^1,\, \E f_k = 0 \Big\} \\
g^*_k(x_k) &= \arg\min_{g_k} \Big \{
   \E \Big( h(X) - g_k(X_k) \Big)^2 \,:\, g_k \in \mh \mathcal{C}^1 ,\, \E g_k = 0 \Big\}
\end{align*}
be the best convex and concave univariate approximation respectively.\\

Then, $f^*_k = 0$ and $g^*_k = 0$ iff $h^*_k(x_k) \equiv \E[ h(X) \given x_k ] = 0$.
\end{proposition}

First, we will establish that $h^*_k(x_k)$ is twice differentiable and that $\partial^2_{x_k} h^*_k(x_k)$ is lower bounded.

\begin{align*}
h^*_k(x_k) &= \E[ h(X) \given x_k] \\
  &= \int_{x_{-k}} h(x) p(\mathbf{x}_{-k} \given x_k) d \mathbf{x}_{-k}\\
\partial^2_{x_k} h^*_k(x_k) 
  &= \int_{\mathbf{x}_{-k}} h''(\mathbf{x}) p(\mathbf{x}_{-k} \given x_k) +
  2 h'(x) p'(\mathbf{x}_{-k} \given x_k) + h(x) p''(\mathbf{x}_{-k} \given x_k) 
  d\mathbf{x}_{-k} 
\end{align*}

The first term $h''(x) p(\mathbf{x}_{-k} \given x_k)$ is strictly positive. By assumption, the remaining terms are continuous and hence bounded on a compact set. $\partial^2_{x_k} h^*_k(x_k)$ is therefore lower bounded. Before proceeding, we also note that because $\E h(X) = 0$, it must be that $\E h^*_k(X_k) = 0$.

Now suppose $f^*_k = 0$ and $g^*_k = 0$. Let $\sigma_k^2$ denote $\E X_k^2$. Then 
\begin{equation}
\arg\min_{c \in \mathbb{R}} \mathbb{E}\Big( h(X) - c (X_k^2-\sigma_k^2) \Big)^2 = 0
\end{equation}
Since optimal $c^* = \frac{\mathbb{E} [ h(X) (X_k^2-\sigma_k^2) ]}{\mathbb{E}[ X_k^2 ]}$, 
we know $\mathbb{E} [ h(X) X_k^2 ] = 
\mathbb{E} \Big[ \mathbb{E}[ h(X) \,|\, X_k] X_k^2 \Big] = 0$.\\

Because $\partial^2_{x_k} h^*_k(x_k)$ is lower bounded, for large enough $\alpha$, $h^*_k(x_k) + \alpha (x_k^2 - \sigma_k^2)$ has a non-negative second derivative and thus is convex. Then
\begin{equation}
\arg\min_{c \in \mathbb{R}} \mathbb{E}\Big( h(X) 
       - c (h^*_k(X_k) + \alpha (X_k^2 - \sigma_k^2)) \Big)^2 = 0
\end{equation}
Again, $c^* = \frac{\mathbb{E}[g(X)\big( 
            h^*_k(X_k) + \alpha (X_k^2 - \sigma_k^2) \big)]}{\mathbb{E}
        \big( h^*_k(X_k) + \alpha (X_k^2 - \sigma_k^2) \big)^2} = 0$, so
\begin{align*}
\mathbb{E}[ h(X) \big( h^*_k(X_k) + \alpha X_k^2 \big) ] &= 
\mathbb{E}[ h(X) h^*_k(X_k) ] \\
& = \mathbb{E}\Big[ \mathbb{E}[h(X) | X_k]  h^*_k(X_k) \Big] \\
& = \mathbb{E} h^*_k(X_k)^2 = 0
\end{align*}
Therefore, $h^*_k(x_k) = 0$.

\end{proof}

>>>>

\begin{example}
\label{examp:gaussian_counterexample}
Consider a quadratic form $f( \mathbf{x}) = \mathbf{x}^\tran H \mathbf{x} + c^\tran \mathbf{x}$ and a Gaussian distribution $X \sim N(0, \Sigma)$ where $\Sigma_{jj} = 1$ for all $j$.
As we show in the appendix, the additive approximation has the
following closed form.
\begin{enumerate}
\item If $f$ does not depend on $x_j$, then $f^*_j(x_j) = 0$.
\item If $f$ depends on $x_j$, then
\begin{equation}
f^*_j(x_j) = H_j^\tran \Sigma_j x_j^2 + c_j x_j
\end{equation}
where $H_j$ and $\Sigma_j$ are the $j$th rows of $H_j$ and $\Sigma_j$, respectively. 
\end{enumerate}
Let $H = \begin{pmatrix}1 & 2 \\ 2 & 5\end{pmatrix}$ and let $\Sigma
= \begin{pmatrix}1 & -\alpha \\ -\alpha & 1\end{pmatrix}$. 
It can be checked that if $\alpha=\frac{1}{2}$, then $f^*_1 = 0$
and additive faithfulness is violated.  Moreover, if $\alpha > \frac{1}{2}$, then $f^*_1$ is a concave function.
\end{example}

@@

Although convex functions are additively faithful---under appropriate conditions---it is difficult to
estimate the optimal additive functions $f^*_k$s as defined in
equation~\eqref{eqn:unconstrained_additive}.  The reason is that $f^*_k$ need not be
a convex function, as example~\ref{examp:gaussian_counterexample}
shows. 
%The function $f^*_k$ can be shown to be smooth but it is preferable to use
%shape-constraints whenever possible to avoid the introduction of the
%additional smoothing bandwidth.
Since the true regression function $f$ is convex, it is natural to ask when it is sufficient to estimate an convex additive model
\begin{equation}
\label{eqn:convex_additive}
\{ f^*_k \}_{k=1}^p = \arg\min \Big \{ 
    \E\Bigl( f(X) - \sum_{k=1}^p f_k(X_k) \Bigr)^2 \,:\, f_k \in \mathcal{C}^1, \, \E f_k(X_k) = 0 \Big \}
\end{equation}
where $\mathcal{C}^1$ is the set of univariate convex functions.
While the convex functions $\{f^*_k\}$ are not additively faithful by
themselves, faithfulness can be restored by coupling the $f^*_k$s with a set of concave functions:
\begin{equation}
\label{eqn:concave_postprocess}
g^*_k = \arg\min \Big\{
   \E\Bigl( f(X) - \sum_{k' \neq k} f^*_{k'}(X_{k'}) - g_k \Bigr)^2
    \,:\, g_k \in \mh \mathcal{C}^1, \E g_k(X_k) = 0 
  \Big\}.
\end{equation}

\begin{theorem}
\label{thm:acdc_faithful}
Suppose $p(\mathbf{x})$ is a positive density on $C=[0,1]^p$ that
satisfies the boundary points condition. Suppose that $f$ is convex
and twice-differentiable, and that that $\partial_{x_k} f$,
$\partial_{x_k} p( \mathbf{x}_{-k} \given x_k )$, and
$\partial_{x_k}^2 p( \mathbf{x}_{-k} \given x_k)$ are all continuous
as functions on $C$.  Let $f^*_k$ and $g^*_k$ be as defined in equations
\eqref{eqn:convex_additive} and \eqref{eqn:concave_postprocess}.  Then
$f^*_k = 0$ and $g^*_k = 0$ implies that $\partial_{x_k} f(\mathbf{x})
= 0$ with probability one,
so that $f$ does not depend on~$x_k$.
\end{theorem}

This theorem naturally suggests a two-stage screening procedure for
variable selection.  In the first stage we fit a sparse convex additive model.
In the second stage, for every variable marked as irrelevant in the first stage,
we fit a univariate \emph{concave} function on the residual for
that variable. We refer to this procedure as AC/DC (additive
convex / decoupled concave) and describe it in more detail in
Section~\ref{sec:acdc}.

\begin{lemma}
\label{cor:faithfulness_extension}
Suppose $p(x)$ is a positive density on $[0,1]^p$ satisfying the boundary points condition.
For any function $\phi(X_{-k})$ that does not depend on $X_k$,
 \begin{align}
f^*_k = \arg\min_{f_k} \mathbb{E} \Bigl[\bigl( f(X) 
           - \phi(X_{-k}) - f_k(X_k) \bigr)^2\Bigr]
\end{align}
is given by $f^*_k(x_k) = \mathbb{E}\bigl[ f(X) - \phi(X_{-k}) \,|\, x_k\bigr]$,
and $ f^*_k = 0$ implies that $\partial_{x_k} f(x) = 0$.
\end{lemma}

\begin{proof}
  In the proof of Theorem~\ref{thm:convex_faithful}, the only property
  of $r(\mathbf{x}_{-k})$ we used was the fact that $\partial_{x_k}
  r(\mathbf{x}_{-k}) = 0$. Therefore, the proof here is identical to
  that of Theorem~\ref{thm:convex_faithful} except that we let
  $\phi(\mathbf{x}_{-k})= r(\mathbf{x}_{-k}).$
\end{proof}

\begin{proof}[Proof of theorem~\ref{thm:acdc_faithful}]
Fix $k$. Let $f^*_k$ and $g^*_k$'s be defined as in
equation~\ref{eqn:convex_additive} and
equation~\ref{eqn:concave_postprocess}.  Then
we have that
\begin{align*}
f^*_k &= \arg\min_{f_k} \Big\{
   \E\big( f(X) - \sum_{k' \neq k} f^*_{k'}(X_{k'}) - f_k \big)^2 
    \,:\, f_k \in  \mathcal{C}^1,\, \E f_k(X_k) = 0 
  \Big\} \\
g^*_k &= \arg\min_{g_k} \Big\{
   \E\big( f(X) - \sum_{k' \neq k} f^*_{k'}(X_{k'}) - g_k \big)^2 
    \,:\, g_k \in \mh \mathcal{C}^1,\, \E g_k(X_k) = 0 
  \Big\}
\end{align*}
Therefore, $f^*_k = 0$ and $g^*_k = 0$ implies that $\E[ f(X) -
\sum_{k'\neq k} f^*_{k'}(X_{k'}) \given x_k] = 0$.  Now we use
Lemma~\ref{cor:faithfulness_extension} with $\phi(\mathbf{x}_{-k}) =
f(\mathbf{x}) - \sum_{k' \neq k} f^*_{k'} (x_{k'})$ and conclude that
$f^*_k = 0$ and $g^*_k = 0$ together imply that $f$ does not depend on $x_k$.
\end{proof}

%%% JL:  I'm commenting out this proposition.  I don't see that it
%% serves any particular role in the rest of the paper.  Am I missing
%% something??
%%
%%\begin{proposition}
%%\label{prop:shape_approx_univariate}
%%Let $C \subset \R^p$ be a compact set and let $h : C \rightarrow \R$. Let $p(x)$ be a positive density on $C$ and suppose $\E h(X) = 0$. Suppose that $\partial_{x_k} h(x)$, $\partial_{x_k} p( x \given x_k)$, and $\partial^2_{x_k} p( x \given x_k )$ are all continuous as functions on $C$. Suppose that $\partial^2_{x_k} h(x) \geq 0$.
%%
%%
%%Let 
%%\begin{align*}
%%f^*_k(x_k) &= \arg\min_{f_k} \Big\{
%%   \E \Big( h(X) - f_k(X_k) \Big)^2 \,:\, f_k \in \mathcal{C}^1,\, \E f_k = 0 \Big\} \\
%%g^*_k(x_k) &= \arg\min_{g_k} \Big \{
%%   \E \Big( h(X) - g_k(X_k) \Big)^2 \,:\, g_k \in \mh \mathcal{C}^1 ,\, \E g_k = 0 \Big\}
%%\end{align*}
%%be the best convex and concave univariate approximation respectively.\\
%%
%%Then, $f^*_k = 0$ and $g^*_k = 0$ iff $h^*_k(x_k) \equiv \E[ h(X) \given x_k ] = 0$.
%%\end{proposition}
%%
%%\begin{proof}
%%
%%
%%First, we will establish that $h^*_k(x_k)$ is twice differentiable and that $\partial^2_{x_k} h^*_k(x_k)$ is lower bounded. 
%%
%%\begin{align*}
%%h^*_k(x_k) &= \E[ h(X) \given x_k] \\
%%  &= \int_{x_{-k}} h(x) p(\mathbf{x}_{-k} \given x_k) d \mathbf{x}_{-k}\\
%%\partial^2_{x_k} h^*_k(x_k) 
%%  &= \int_{\mathbf{x}_{-k}} h''(\mathbf{x}) p(\mathbf{x}_{-k} \given x_k) +
%%  2 h'(x) p'(\mathbf{x}_{-k} \given x_k) + h(x) p''(\mathbf{x}_{-k} \given x_k) 
%%  d\mathbf{x}_{-k} 
%%\end{align*}
%%
%%The first term $h''(x) p(\mathbf{x}_{-k} \given x_k)$ is strictly positive. By assumption, the remaining terms are continuous and hence bounded on a compact set. $\partial^2_{x_k} h^*_k(x_k)$ is therefore lower bounded. Before proceeding, we also note that because $\E h(X) = 0$, it must be that $\E h^*_k(X_k) = 0$.
%%
%%Now suppose $f^*_k = 0$ and $g^*_k = 0$. Let $\sigma_k^2$ denote $\E X_k^2$. Then 
%%\begin{equation}
%%\arg\min_{c \in \mathbb{R}} \mathbb{E}\Big( h(X) - c (X_k^2-\sigma_k^2) \Big)^2 = 0
%%\end{equation}
%%Since optimal $c^* = \frac{\mathbb{E} [ h(X) (X_k^2-\sigma_k^2) ]}{\mathbb{E}[ X_k^2 ]}$, 
%%we know $\mathbb{E} [ h(X) X_k^2 ] = 
%%\mathbb{E} \Big[ \mathbb{E}[ h(X) \,|\, X_k] X_k^2 \Big] = 0$.\\
%%
%%Because $\partial^2_{x_k} h^*_k(x_k)$ is lower bounded, for large enough $\alpha$, $h^*_k(x_k) + \alpha (x_k^2 - \sigma_k^2)$ has a non-negative second derivative and thus is convex. Then
%%\begin{equation}
%%\arg\min_{c \in \mathbb{R}} \mathbb{E}\Big( h(X) 
%%       - c (h^*_k(X_k) + \alpha (X_k^2 - \sigma_k^2)) \Big)^2 = 0
%%\end{equation}
%%Again, $c^* = \frac{\mathbb{E}[g(X)\big( 
%%            h^*_k(X_k) + \alpha (X_k^2 - \sigma_k^2) \big)]}{\mathbb{E}
%%        \big( h^*_k(X_k) + \alpha (X_k^2 - \sigma_k^2) \big)^2} = 0$, so
%%\begin{align*}
%%\mathbb{E}[ h(X) \big( h^*_k(X_k) + \alpha X_k^2 \big) ] &= 
%%\mathbb{E}[ h(X) h^*_k(X_k) ] \\
%%& = \mathbb{E}\Big[ \mathbb{E}[h(X) | X_k]  h^*_k(X_k) \Big] \\
%%& = \mathbb{E} h^*_k(X_k)^2 = 0
%%\end{align*}
%%Therefore, $h^*_k(x_k) = 0$.
%%
%%\end{proof}
%%
